---
title: "Class 3"
---


```{r setup, include=FALSE}
# Global knitr options
knitr::opts_chunk$set(
  warning = FALSE, # Suppresses warnings
  message = FALSE  # Suppresses messages
)
```

# Objectives

* To get some practice using the tools we have discussed so far.

# Goal

The goal will be to analyze some insurance claims data. Install the `RESI` `R` package load the package and the data.

```{r, eval=FALSE}
install.packages('RESI')
```

This code below loads the `insurance` dataset and assigns it to the variable `ins`.

```{r}
library(RESI)
ins = insurance
head(ins)
```

We'll follow the outline of a paper:

1. **Introduction** -- this part is roughly done for you.
2. **Methods**
    - Convert the research questions into conclusions you can make about parameters in a linear regression model.
    - Table 1. What variables will you include? What tests will you perform in the table?
    - What exploratory visualizations will you use?
    - What model will you use to answer the research questions. Do you need to construct a test, confidence interval, or both?
    - What visualizations will you use to evaluate the model assumptions?
    - How will you visualize the results with the data.
3. **Results**
    - Follow the structure we outlined in the Methods and perform the analyses. For a report like this, you can include everything. For a paper, you will work with the team to decide what information is most important to include in the document.
4. **Discussion**
    - We won't worry about this for now

# Introduction

Let's use the insurance dataset to investigate the following questions.

1. **Age Influence on Health Charges:**
   - How does age affect medical charges, and is the relationship linear, or do older age groups experience disproportionately higher charges?
2. **Influence of Smoking on Medical Expenses:**
   - How does smoking status influence medical charges, and are smokers subjected to significantly higher health-related expenses compared to non-smokers?
3. **Regional Variations in Healthcare Charges:**
   - Are there regional differences in medical charges, and how do regional factors potentially contribute to the variation in healthcare costs?
4. **Combined Effects of Predictors:**
   - What does the relationship between each variable look like after controlling for the others? **Hint:** Use added variable plots.
   - Does smoking status affect the influence of age on medical charges?


# Methods

**Describe the overall strategy:**

For all study aims, we will use a `XX` model including `XX` as covariates.
We will use `XX` plots to understand the distributions of the covariates with the outcome.
We will evaluate the effect of each variable using XX tests and include added variable plots to visualize the unique contribution of each variable.
**Other ideas?**


Then describe the specific strategy for each research question. Create the results below.

1. **Age Influence on Health Charges:**
   - To evaluate the effect of age on medical charges, we will test the linear age effect using a `XX` test and report the effect with the estimate, confidence interval, T-statistic, degrees of freedom, and $p$-value.
   - To evaluate the linearity of the relationship, we will do `XX`.
   - Testing `XX` will evaluate if older ages experience higher charges.
2. **Influence of Smoking on Medical Expenses:**
   - *fill in your answer here*
3. **Regional Variations in Healthcare Charges:**
   - *fill in your answer here*
4. **Combined Effects of Predictors:**
   - *fill in your answer here*


# Results

**Follow the same structure as methods, presenting the results:**

Describe the specific strategy for each research question and perform the analyses/create the results below.


## Add Table 1:

*fill in your answer here*

## Add the exploratory analyses here:

*fill in your answer here*

## Age Influence on Health Charges:

To evaluate the effect of age on medical charges, perform a test of the linear age effect using a `XX` test and report the effect with the estimate, confidence interval, T-statistic, degrees of freedom, and $p$-value.

## Influence of Smoking on Medical Expenses:

*fill in your answer here*


## Regional Variations in Healthcare Charges:

*fill in your answer here*

## **Combined Effects of Predictors:**
   
*fill in your answer here*



# Discussion


