---
title: "Class 3"
format:
  html:
    theme: cosmo
    css: styles.css
    toc: true
    code-fold: true
---


```{r setup, include=FALSE}
# Global knitr options
knitr::opts_chunk$set(
  warning = FALSE, # Suppresses warnings
  message = FALSE  # Suppresses messages
)
```

# Objectives

* To get some practice using the tools we have discussed so far.

# Goal

The goal will be to analyze some insurance claims data. Install the `RESI` `R` package load the package and the data.

```{r, eval=FALSE}
install.packages('RESI')
```

This code below loads the `insurance` dataset and assigns it to the variable `ins`.

```{r}
library(RESI)
ins = insurance
head(ins)
```

We'll follow the outline of a paper:

1. **Introduction** -- this part is roughly done for you.
2. **Methods**
    - Convert the research questions into conclusions you can make about parameters in a linear regression model.
    - Table 1. What variables will you include? What tests will you perform in the table?
    - What exploratory visualizations will you use?
    - What model will you use to answer the research questions. Do you need to construct a test, confidence interval, or both?
    - What visualizations will you use to evaluate the model assumptions?
    - How will you visualize the results with the data.
3. **Results**
    - Follow the structure we outlined in the Methods and perform the analyses. For a report like this, you can include everything. For a paper, you will work with the team to decide what information is most important to include in the document.
4. **Discussion**
    - We won't worry about this for now

# Introduction

Let's use the insurance dataset to investigate the following questions.

1. **Age Influence on Health Charges:**
   - How does age affect medical charges? is the relationship linear or nonlinear?
2. **Influence of Smoking on Medical Expenses:**
   - How does smoking status influence medical charges, and are smokers subjected to significantly higher health-related expenses compared to non-smokers?
3. **Does smoking affect the charges age trajectory:** Is there evidence that smoking changes the effect of age on charges?
4. **Regional Variations in Healthcare Charges:**
   - Are there regional differences in medical charges, and how do regional factors potentially contribute to the variation in healthcare costs?
5. **BMI:**
   - Does BMI relate to insurance charges? Is it nonlinear?
6. **Combined Effects of Predictors:**
   - What does the relationship between each variable look like after controlling for the others? **Hint:** Use added variable plots.
   - Does smoking status affect the influence of age on medical charges?

# Methods

**Describe the overall strategy:**

* For all study aims, use charges as the outcome including age, sex, region, bmi, and smoking status as covariates (independent variables).
* **We want to know about the effect of each variable controlling for the others.**
* Use histograms, boxplots, scatter plots, and Table 1 to understand the distributions of the outcome and the relationship between the outcome and covariates.
* Evaluate the effect of each variable using T-tests, F-tests. Use a single model, or a sequence of model comparisons.
* You can use added variable plots to visualize the unique contribution of each variable.

**Variables to include in Table 1:**

* Charges
* Age
* Sex
* BMI
* Region
* Smoking status


**Consider using these tools**

* Regression diagnostic plots.
* Transforming the outcome (e.g. `log` transformation)
* Allowing for nonlinearity of continuous terms.
* Robust standard errors (in the `car` package, `Ancova(model, vcov.=vcovHC, type=2)`)

Then describe the specific strategy for each research question. Create the results below.

1. **Age Influence on Health Charges:**
   - To evaluate the effect of age on medical charges, we will test the linear age effect using a T-test and report the effect with the estimate, confidence interval, T-statistic, degrees of freedom, and $p$-value.
   - To evaluate the linearity of the relationship, we will test for a nonlinear effect over the linear effect.
2. **Influence of Smoking on Medical Expenses:**
   - To evaluate the effect of smoking status on medical charges, we will ...
   - Testing the coefficient for smoking.
3. **Smoking effect on age effect**
  - To evaluate the effect of smoking on the age effect, we ...
  - To understand how the age effect differs, we used `emmeans`
4. **Regional Variations in Healthcare Charges:**
   - We will perform an ____ to assess regional differences in charges.
   - To understand how regions differ, we create a plot (use `emmeans`)
5. **BMI:**
6. **Combined Effects of Predictors:**
   - Added variable plots.
   - $R^2$


# Results

**Follow the same structure as methods, presenting the results:**

Describe the specific strategy for each research question and perform the analyses/create the results below.


## Add Table 1:

```{r}
library(table1)
table1(~charges + age + sex + region + bmi | smoker, data=ins, caption='Demographics stratified by smoking status.' )
```

## Add the exploratory analyses here:


```{r}
hist(ins$charges, main='Histogram of insurance charges', xlab='Charges')
```



### Evaluate model fit

Does anything need to change about the model you fit?

## Model results

It usually makes sense to report all coefficients, tests, confidence intervals, degrees of freedom, and $p$-values in a table and refer to the table.

* Create a "summary" table with all coefficients in the model.
* Create an anova table with type 2 sum of squares testing all effects in the model.


## Age Influence on Health Charges:

To evaluate the effect of age on medical charges, we performed a test of the linear age effect using a T-test and report the effect with the estimate, confidence interval, T-statistic, degrees of freedom, and $p$-value.

There was evidence that age was positive associated with medical charges (b=258.54, StdEr=11.94, $T=21.66$, $df=1300$, $p <0.001$).

## Influence of Smoking on Medical Expenses:

*Interpret the smoking effect here*


## Smoking effect on age association

*Interpret the smoking effect on the age effect here*

## Regional Variations in Healthcare Charges:

*Interpret the regional differences in healthcare charges here*



# Discussion


