---
title: "Class 4: Logistic regression"
editor: visual
---

# Objectives

* Interpret logistic regression coefficients.
* Evaluate

# Logistic regression



```{r}
library(RESI)
ins = insurance

glmModel = glm(smoker ~ age + sex + region, data=ins, family=binomial())
summary(glmModel)
```



# Interpreting Logistic Regression Interpretation


## Model
   Logistic regression model predicts the log odds of the binary outcome (e.g., whether an individual is a smoker or not) as a linear combination of the predictor variables.

   The model can be formulated as:
   $$
   \text{logit}(p) = \log\left(\frac{p}{1 - p}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \ldots + \beta_k X_k
   $$
   where:
   - $ p $ is the probability of the outcome (smoking = 1).
   - $ X_1, X_2, \ldots, X_k $ are the predictor variables (age, sex, region).
   - $ \beta_0 $ is the intercept.
   - $ \beta_1, \beta_2, \ldots, \beta_k $ are the coefficients for the predictors.

## Intepreting coefficients on the log-odds scale
   Each coefficient $ \beta_i $ represents the change in the log odds of the outcome for a one-unit increase in the corresponding predictor variable $ X_i $, holding all other predictors constant. 

   - **For continuous predictors (e.g., Age)**:
     - If $ \beta_1 =-0.004$ (age): for every 1-year increase in age, the log odds of being a smoker increase by 0.05.

   - **For categorical predictors (e.g., Sex and Region)**:
     - If $ \beta_2 = 0.37$, then being male decreases the log odds of being a smoker by 1.5.

## Interpreting coefficients on the odds scale
   - To convert the log odds to odds, we exponentiate:
   \[
   \text{odds} = e^{\text{logit}(p)} = e^{\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \ldots + \beta_k X_k}
   \]

   - **Odds Ratio (OR)**:
     - The odds ratio for a one-unit increase in $ X_i $ is given by:
     \[
     \text{OR} = e^{\beta_i}
     \]
     - This tells us how the odds of the outcome change with a one-unit increase in $ X_i $:
       - If $ \text{OR} > 1 $, the odds of being a smoker increase (more likely).
       - If $ \text{OR} < 1 $, the odds of being a smoker decrease (less likely).
       - If $ \text{OR} = 1 $, there is no change in the odds.



## Example visualization

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Create a new dataframe for predictions
ins$predicted_prob = predict(glmModel, type = "link")

# Create a visualization of predicted probabilities against age
ggplot(ins, aes(x = age, y = predicted_prob, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  labs(title = "Predicted Probability of Smoking by Age",
       x = "Age",
       y = "Predicted Probability of Smoking") +
  theme_minimal()
```



## Robust standard errors

I like to use robust standard errors for testing whenever I can. They are also possible for the `glm` function.

```{r}
library(sandwich) 
library(lmtest)

coeftest(glmModel, vcov = vcovHC(glmModel))

car::Anova(glmModel, vcov = vcovHC(glmModel))
```
